Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-04 doc/paper.Rmd
fixed some references

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	# fig.show = "hide" #hide all figures until the end
	dpi = 300
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

@@

Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @brunaExperimentalAssessmentHeliconia2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].

@@

```{r eda, fig.cap=eda_cap, fig.height=7, fig.width=10}
eda_plot
```

@@

```{r size-plot, fig.cap=size_cap, fig.height=8, fig.width=5}
size_plot
```

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].

@@

Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@laurance2014].
However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	fig.show = "hide", #don't knit figures for now.
	dpi = 300
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

@@

Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @brunaExperimentalAssessmentHeliconia2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].

@@

```{r eda, fig.cap=eda_cap}
eda_plot
```

@@

```{r size-plot, fig.cap=size_cap, fig.height=6, fig.width=5}
size_plot
```

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].

@@

Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017)
